---
title: "Assignment1 - Eli Lauwers"
output: 
  rmarkdown::github_document:
    number_sections: true
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library("DBI")
library("readr")
library("jsonlite")

knitr::opts_chunk$set(
  echo = TRUE
)
database_creds = read_json("database_creds.json")
database_creds[["drv"]] = RPostgres::Postgres()

con = do.call(
  RPostgres::dbConnect, 
  database_creds
)
printlines = 25
```


# Consider the relational schema of the `rollsbin` database

## Is it possible to assign multiple (different) locations to the same branch? Why (not)?

This question is a bit dependent on what exactly is a branch and how it is related to an enterprise(name). I see two options, and I will give an explanation for both.

(1) If two different numbers of `branch.enterprisenumber` are considered two different branches, than the answer is 'no, multiple locations cannot be assigned to the same location'. Reason being that `branch.enterprisenumber`is the primary key, and will always be unique in the `branch` table. That means that a given number can only appear once in the table, and every enterprise in that sense will be coupled with one -and only one- location.

(2) If two different names in `branch.name` are considered two different branches, then the answer is 'yes, multiple locations can be assigned to one branch'. There is no constraint on the `branch.name` column, so two rows might exist with identical `branch.name`. Every enterprise -or row- in the `branch` table has a location given by `branch.postalcode` and `branch.municipality`. If there are two rows with the same name and different locations, then we have a situation where multiple locations are assigned to the same branch.

I consider (2) to be the most logical case for this exercise. My final answer follows (2)

**Answer**: Yes, multiple locations can be assigned to the same branch.

## Is it possible to store locations that are not linked to a branch? Why (not)?

Yes, the only constraint for the `location`-table is that the pair of `location.postalcode` and `location.muncipality`, being the (combined) primary key must be unique. There is however no constraint that every 'location' must be present somewhere in the `branch` table.

Next, we know from the relational schema that locations are stored for the branches as well as for the employee working for those branches. If every location should have to be linked to a branch, than branch employees can only live in the location of a branch, which would be quite stupid.

**Answer**: Yes, it is possible to store locations not linked to a branch.

# Suppose one wants to select all data 

Suppose one wants to select all data (i.e. data stored in columns license_plate, enterprisenumber, brand and model) related to cars of type ‘cabriolet’. What is a potential issue that one may have when executing the following SELECT-query for this task and why? How would you solve this? Explain in your own words.

```{sql connection = con, eval = F}
SELECT DISTINCT 
  c.license_plate, c.enterprisenumber, c.brand, c.model
FROM car c
  INNER JOIN carmodel cm ON c.brand = cm.brand
  INNER JOIN type t ON cm.type = t.name
WHERE t.name = 'cabriolet'
```

I see three issues:

1. The `type` column only adds a cars price per day to the car information. If a car's price per day is not strictly necessary (rereading the question, this seems to be the case), then the `type` table does not need to be integrated in the answer. The join on `t.name` can be removed.
2. The `WHERE` statement will compare every car name against a fixed string. There is however no additional constraint on the column that vluaes should be only lowercase. A better way of comparing would be to compare against a string without case sensitivity. One can do this by replacing the `=` operator in the `WHERE` clause with an `ilike` clause.
3. I see a problem with the joins. See below.

In the first `INNER JOIN`, we're joining solely on a cars brand. However, in the `carmodel`-table, one brand can be linked with multiple distinct models. Imagine a car of the brand 'FIAT' in the `cars`-table . Next, imagine two carmodels in the `carmodel`-table for the FIAT brand: FIAT **Punto** and FIAT **500**. As the inner join solely joins on brand, both fiat models will be joined to only one car.

If this assumption is true, than in the result of the query we will see identical numberplates linked to multiple distinct carmodels, which shouldn't be possible. I looked around and quickly found proof for this hypthesis:

```{sql connection = con}
SELECT *
FROM car c
  INNER JOIN carmodel cm ON c.brand = cm.brand
  INNER JOIN type t ON cm.type = t.name
WHERE t.name = 'cabriolet' AND c.license_plate = '1-IZX-389'
```

You can see that one license plate is linked to multiple car types, which should not be possible.

To improve the query, one small fix has to be done. In the first `INNER JOIN`, we will join on a cars brand, but also on a cars model. We will then preform the same testquery to see if this strategy works better. I also use the `USING` keyword as a preferation for working with join-statements.

```{sql connection = con}
SELECT *
FROM car c
  INNER JOIN carmodel cm USING(brand, model)
  INNER JOIN type t ON cm.type = t.name
WHERE t.name = 'cabriolet' AND c.license_plate = '1-IZX-389'
```

Looking at the test-query, we see that the problematic situation encountered earlier was overcome!

# Consider the following `SELECT`-query

```{sql connection = con, eval = FALSE}
SELECT * FROM contract c1
  LEFT JOIN contract c2 ON c1.employeenumber = c2.employeenumber
    AND c1.enterprisenumber = c2.enterprisenumber
    AND c1.period_begin < c2.period_begin
WHERE c2.period_begin IS NULL;
```

## Explain, in your own words, what this SELECT-query achieves. 
You should not give the result table of this query or explain this query in technical terms, but explain what the semantical outcome is of this query when executed on data that is stored in the rollsrobin database.

The asterisk sign at the start of the query means that every column from the resulting table will be selected and that no columns will be removed.

Next, a `LEFT JOIN` is performed on the `contract`-table with itself with three join-conditions, where every condition must be met to perform a join:

- `c1.employeenumber = c2.employeenumber`: every employee-contract is linked to every contract for that employee. If an employee has 5 contracts, than the resulting table will have 25 rows (5 * 5 = 25) for that employee.
- `c1.enterprisenumber = c2.enterprisenumber`: An employees contracts will be linked, but only contracts within the same enterprise will be linked
- `c1.period_begin < c2.period_begin`: The left contract can only be linked if the left starts earlier in time than the right one

Next, we filter the resulting table for rows where the `begin_period` for the right contract is `null`.

The resulting table is thus an overview for every employee that shows **the last** contract for a given enterprise. If an employee has worked for multiple enterprises, than there will be a row for every enterprise the employee has worked for.

## Draw the table

Draw the entire table that will be returned upon completion of the given query, starting from the example data of the contract table, given in the appendix of this document.

| employeenumber | period_begin | period_end | enterprisenumber | employeenumber | period_begin | period_end | enterprisenumber |
|----------------|--------------|------------|------------------|----------------|--------------|------------|------------------|
| 1              | 01/07/2021   | 31/12/2021 | 102              | null           | null         | null       | null             |
| 1              | 01/08/2021   | 01/08/2023 | 101              | null           | null         | null       | null             |
| 2              | 24/02/2022   | 15/05/2022 | 101              | null           | null         | null       | null             |
| 2              | 31/05/2022   | 30/06/2022 | 104              | null           | null         | null       | null             |
| 3              | 01/10/2021   | 30/09/2022 | 103              | null           | null         | null       | null             |

## Draw again

Suppose one removes the entire WHERE-clause from the SELECT-query.
Draw, again, the entire table that will be returned upon completion of
the given query (now without the WHERE-clause), starting from the example
data of the contract table, given in the appendix of this document.

| employeenumber | period_begin | period_end | enterprisenumber | employeenumber | period_begin | period_end | enterprisenumber |
|----------------|--------------|------------|------------------|----------------|--------------|------------|------------------|
| 1              | 12/01/2021   | 31/05/2021 | 101              | 1              | 01/02/2022   | 30/06/2022 | 101              |
| 1              | 12/01/2021   | 31/05/2021 | 101              | 1              | 01/08/2022   | 01/08/2023 | 101              |
| 1              | 01/02/2022   | 30/06/2022 | 101              | 1              | 01/08/2022   | 01/08/2023 | 101              |
| 1              | 01/07/2021   | 31/12/2021 | 102              | null           | null         | null       | null             |
| 1              | 01/08/2021   | 01/08/2023 | 101              | null           | null         | null       | null             |
| 2              | 24/02/2022   | 15/05/2022 | 101              | null           | null         | null       | null             |
| 2              | 31/05/2022   | 30/06/2022 | 104              | null           | null         | null       | null             |
| 3              | 01/10/2021   | 30/09/2022 | 103              | null           | null         | null       | null             |

## Add a row

Is it possible to add the row {employeenumber: 1, period_begin: 12/01/2021, period_end: 31/01/2021, enterprisenumber: 103} to the contract table given in the appendix of this document, considering the relational schema of the rollsrobin database and the example data? Why (not)?

Yes, this is possible. For the `contracts`-table, the primary key is a combined key for the `contract.employeenumber`, `contract_begin` and `contract.period_end`, which means that the combination of those elements must be unique. There is a row where the employeenumber and begin date are equal to this row of interest, but the period ends differ. 

Next, there is a constraint on the `contract`-table that a new row must always have a begin date that is prior to the end date. That condition is met in this test row

**Answer**: Yes, we can safely add the new row

# Imagine

Imagine that the designers of the rollsrobin database would have left out the employee table from the relational schema. Instead, all data related to employees would then be stored in the person table (which would include an extra column employeenumber). The foreign key in the table contract, originally pointing to the employee table, would then point to the column employeenumber in the person table. What are possible (dis)advantages of this approach? Explain in your own words.

Disadvantage: the `employeenumber` in the `person`-table must become optional, as basic customers will not have an employeenumber. In that sense, it is crucial that an employee can only be made with an employeenumber, which is a check that has to be done by the form which creates new employees.

Disadvantage: As of now, personal information that is stored for regular persons / customers is the same as personal information stored for the employees. If in the future there should come a company policy to store more information on employees, than the `employee` table is the best place to store that, as every employee has personal information in the `person` table, with additional employee information in the `employee` table.

Advantage: The only added value of the `employee` table is the link from an emailadress to an employeenumber. It might be a bit excessive to create and maintain an entire table for one piece of information. By adding one column to the `person` table, a whole table can be removed.

 
**Answer**: Although the `employee` table can be removed with the addition of one column to the `person` table, I would advise against integrating  the `employee` and `person` tables. 

# Basic SQL

## Postal codes and month letters

Construct a list of unique email addresses of all persons who did a registration for which (1) the rented car is owned by a branch for which the sum of the numbers in the branches postal code (branch.postalcode) equals the sum of the numbers in the postal code of the person, who rented the car (person.postalcode), and (2) the first date of the rental period (registration.period_begin) is in a month which name ends with letter ‘r’ (in English). You may assume that all postal codes consist of 4 digits between 0 and 9. In the result table, we expect one column with name email of datatype varchar.

```{sql first exercise,code=readLines("01504786_Eli_Lauwers_1.sql"),connection=con,eval=F}

```

```{r,echo=FALSE}
df1 = dbGetQuery(con, statement = read_file("01504786_Eli_Lauwers_1.sql"))
write.csv(df1,"csv_tables/df1.csv")
knitr::kable(head(df1,printlines),format="markdown")
```

## Rentals during contract time

Construct a list of unique email addresses of all employees who rented a car at a company at which they have worked. Moreover, the first date of the rental period should fall within the period they were working at the company (i.e. registration.period_begin should be between contract.period_begin and contract.period_end, boundaries inclusive). In the result table, we expect one column with name email of datatype varchar. Order the results in this table alphabetically according to the values in the column email. 


```{sql second exercise,code=readLines("01504786_Eli_Lauwers_2.sql"),connection=con,eval=F}

```

```{r,echo=FALSE}
df2 = dbGetQuery(con, statement = read_file("01504786_Eli_Lauwers_2.sql"))
write.csv(df2,"csv_tables/df2.csv")
knitr::kable(head(df2,printlines),format="markdown")
```

## Unregistered locations

Construct a list of locations (postal code and municipality) in which each location (1) is not registered as a residence of a person in the database, (2) has an even postal code and (3) does not start with letter ‘B’ (case insensitive). In the result table, we expect two columns with corresponding datatype: postalcode (varchar) and municipality (varchar).

```{sql third exercise,code=readLines("01504786_Eli_Lauwers_3.sql"),connection=con,eval=F}

```

```{r,echo=FALSE}
df3 = dbGetQuery(con, statement = read_file("01504786_Eli_Lauwers_3.sql"))
write.csv(df3,"csv_tables/df3.csv")
knitr::kable(head(df3,printlines),format="markdown")
```

## Number of parts in municipality

Give for each location in the database (uniquely defined by postalcode and municipality), the number of parts that make up the name of the municipality. Here, a part is defined as a substring of the municipality that is separated from other substrings by either a hyphen (“-”) or a space (“ ”).

```{sql fourth exercise,code=readLines("01504786_Eli_Lauwers_4.sql"),connection=con,eval=F}

```

```{r,echo=FALSE}
df4 = dbGetQuery(con, statement = read_file("01504786_Eli_Lauwers_4.sql"))
write.csv(df4,"csv_tables/df4.csv")
knitr::kable(head(df4,printlines),format="markdown")
```

## Customer pairs

Give all possible couples of registrations that were done by the same person (i.e. the table that is returned by your query should contain rows consisting of two registrations that were conducted by the same person). Be aware of the fact that no rows may be present in your result that contain two identical registrations (i.e. same person, car, begin date and end date)! In order to prevent this, the value in column period_begin of the first registration should be chronologically strict before the value in column period_begin of the second registration. However, if this value is the same for the two registrations, the value in column license_plate of the first registration should come alphabetically strict before the value in column license_plate of the second registration.

```{sql fifth exercise,code=readLines("01504786_Eli_Lauwers_5.sql"),connection=con,eval=F}

```

```{r,echo=FALSE}
df5 = dbGetQuery(con, statement = read_file("01504786_Eli_Lauwers_5.sql"))
write.csv(df5,"csv_tables/df5.csv")
knitr::kable(head(df5,printlines),format="markdown")
```